---
title: 'Reproducible Research: Peer Assessment 2'
author: "John R Ferguson"
date: "December 31, 2018"
output: html_document
---

#### Below the top 15 categories of fatalities, injuries and economic impact to crops and property is shown.  Tornados kill and injure the most and High winds, Heavy rains, and Floods cause the most economic damage.

```{r start}
library(ggplot2)
library(gridExtra)
library(plyr)

setwd("~/coursera")

## Data Processing

#### The data event types needed cleaning.  I cleaned enough to provide good visibility of the top fifteen event types for fatalities and injuries.  Bs, Ms, Ks and Hs had to be replaced with 9s, 6s, 3s, and 2s respectively to perferm economic calculations. The economic top fifteen was easy to aggregate.

```{r init}

projectdata <- read.csv('repdata_data_StormData.csv')

summary(projectdata)

sum(projectdata$FATALITIES)

fdata <- ddply(projectdata,c("EVTYPE","FATALITIES","INJURIES","PROPDMGEXP","CROPDMGEXP"),summarise, F_sum=sum(FATALITIES),I_sum=sum(INJURIES),P_sum=sum(PROPDMG),C_sum=sum(CROPDMG))

fdata_Corr <- fdata
fdata_Corr$EVTYPE <- gsub("^WINTER STORM.*", "WINTER STORM",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^(EXTREME|RECORD/EXCESSIVE|RECORD) HEAT$", "EXCESSIVE HEAT",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^EXCESSIVE HEAT*", "HEAT",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^HEAT WAVES?$|^UNSEASONABLY WARM$|^UNSEASONABLY WARM AND DRY$|^WARM WEATHER$", "HEAT",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^TORNADO.*", "TORNADO",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^FLOOD*|FLASH FLOOD", "FLOOD",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("TSTM|THUNDERSTORMS?|THUNDERSTORM WIND.*|THUNDERSTORM.*", "THUNDERSTORM",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("^ICE.*|BLIZZARD|HEAVY SNOW|HAIL", "WINTER STORM",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("EXTREME COLD/WIND CHILL", "EXTREME COLD",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("THUNDERSTORM WIND", "THUNDERSTORM",fdata_Corr$EVTYPE)
fdata_Corr$EVTYPE <- gsub("WILD/FOREST FIRE", "WILDFIRE",fdata_Corr$EVTYPE)

fdata_Corr$PROPDMGEXP <- gsub("B|b", "9",fdata_Corr$PROPDMGEXP)
fdata_Corr$CROPDMGEXP <- gsub("B|b", "9",fdata_Corr$CROPDMGEXP)
fdata_Corr$PROPDMGEXP <- gsub("M|m", "6",fdata_Corr$PROPDMGEXP)
fdata_Corr$CROPDMGEXP <- gsub("M|m", "6",fdata_Corr$CROPDMGEXP)
fdata_Corr$PROPDMGEXP <- gsub("K|k", "3",fdata_Corr$PROPDMGEXP)
fdata_Corr$CROPDMGEXP <- gsub("K|k", "3",fdata_Corr$CROPDMGEXP)
fdata_Corr$PROPDMGEXP <- gsub("H|h", "2",fdata_Corr$PROPDMGEXP)
fdata_Corr$CROPDMGEXP <- gsub("H|h", "2",fdata_Corr$CROPDMGEXP)

```

## Results

```{r, echo=FALSE}
fdata2 <- ddply(fdata_Corr,c("EVTYPE"),summarise, F_sum=sum(FATALITIES))
fdata2 <- arrange(fdata2,fdata2[,2], decreasing=T)
fdata2 <- head(fdata2,n=15)

idata2 <- ddply(fdata_Corr,c("EVTYPE"),summarise, I_sum=sum(INJURIES))
idata2 <- arrange(idata2,idata2[,2], decreasing=T)
idata2 <- head(idata2,n=15)

edata2 <- ddply(fdata_Corr,c("EVTYPE","PROPDMGEXP","CROPDMGEXP"),summarise, P_sum=sum(P_sum),C_sum=sum(C_sum))

edata2$total <- edata2$P_sum*10^(as.numeric(edata2$PROPDMGEXP))+edata2$C_sum*10^(as.numeric(edata2$CROPDMGEXP))
edata2$ptotal <- edata2$P_sum*10^(as.numeric(edata2$PROPDMGEXP))
edata2$ctotal <- edata2$C_sum*10^(as.numeric(edata2$CROPDMGEXP))

edata3 <- ddply(edata2,c("EVTYPE"),summarise, T_sum=sum(total))

edata3 <- arrange(edata3,edata3[,2], decreasing=T)
edata3 <- head(edata3,n=15)

edata4 <- ddply(edata2,c("EVTYPE"),summarise, P_sum=sum(ptotal))

edata4 <- arrange(edata4,edata4[,2], decreasing=T)
edata4 <- head(edata4,n=15)

edata5 <- ddply(edata2,c("EVTYPE"),summarise, C_sum=sum(ctotal))

edata5 <- arrange(edata5,edata5[,2], decreasing=T)
edata5 <- head(edata5,n=15)
```

### Which types of events are most harmful to population health?

#### Based on the following plots Tornados, Heat, Floods and Winter Storms produce the most fatalities and injuries in that order.

```{r, echo=FALSE}
qplot(EVTYPE, data = fdata2, weight = F_sum, geom = "bar") + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
qplot(EVTYPE, data = idata2, weight = I_sum, geom = "bar") + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
```

### Which types of events have the greatest economic consequences?

#### Based on the following plots High winds, Heavy rains, Floods and Fires produce the most Economic losses in that order.


```{r, echo=FALSE}
qplot(EVTYPE, data = edata3, weight = T_sum, geom = "bar") + 
    scale_y_continuous("Economic Impact") + 
    theme(axis.text.x = element_text(angle = 45, 
                                     hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Economic Impact by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
```


